---
title: "Explorative Data Analysis - control"
format: html
date: "2026-01-20"
execute:
  fig-format: png
  fig.dpi: 300
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(root.dir = "/home/mj/project1-ctrl-stat/eda/scripts")
```

Built with R version `r getRversion()`

# Data preparation

## Environment cleaning and set-up

```{r}
#| output: false
#| message: false
# Set working directory
# setwd("/media/mj/ANTIX-LIVE/project1-ctrl-stat/scripts")

rm(list=ls())
```


```{r}
#| eval: false
#| output: false
#| message: false
# Clean working directory, if needed
rm(list=ls())
# Install required packages
## From Bioconductor
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GenomicRanges")
BiocManager::install("DESeq2")
BiocManager::install("genefilter")
BiocManager::install("apeglm")

## From CRAN
install.packages("pheatmap")
install.packages("RColorBrewer")
install.packages("viridis")
install.packages("PoiClaClu")
install.packages("glmpca")
install.packages("ggplot2")
install.packages("dplyr")
```

## Load data

Loading of count data and metadata.
```{r data-loading, cache=TRUE}
#| output: false
#| message: false

# previously prepared range data
library("GenomicRanges")
load("../../../project1-stat/input-data/rdata/ranges.RDa")

# count data
cts <- as.matrix(read.table("../../input-data/counts/counts-ctrl4r", sep="\t", header=T, row.names="Geneid"))
# check
dim(cts)
cts[1:5, 1:5]

# read-in sample data
coldata=read.table("../../input-data/counts/coldata-ctrl4r", sep="\t", header=T, row.names=1)
#check
dim(coldata)
colnames(coldata)
class(coldata)
coldata[1:5,]
str(coldata)

# bin Million of reads
coldata$Mread_bin <- cut(coldata$Mreads, breaks = 4)  # 4 bins

## make factors
library(dplyr)

coldata <- coldata |>
  mutate(across(c(letter, tissue, line, time, day, rep, Mread_bin), as.factor))
str(coldata)

# Check if files are in concert
all(rownames(coldata) %in% colnames(cts))
all(rownames(coldata) == colnames(cts))

# Order is different - reorder count file
cts <- cts[, rownames(coldata)]

# Check one more time
all(rownames(coldata) %in% colnames(cts))
all(rownames(coldata) == colnames(cts))

# Save coldata and counts to R file for reuse
save(coldata, cts, file="../../input-data/rdata/counts_and_metadata.RDa")
```

## Create DEseq Data Set

```{r dds, cache=TRUE}
#| output: false
#| message: false
library("DESeq2")
# create dataset, design included only for command completeness, different will be used
dds=DESeqDataSetFromMatrix(countData=cts, colData=coldata, rowRanges=gzakresy, design= ~ tissue+line+time)
```

## Pre-filtering

```{r filter, cache=TRUE}
# keeping genes having at least 3 samples with a count of 10 or higher
keep <- rowSums(counts(dds) >= 10) >= 3
dds.f <- dds[keep,]
dim(dds.f)

# what percent of genes remained?
100*(dim(dds.f)[1]/dim(dds)[1])
```

::: {.callout-important title="Advice from Peter Langfelder (WGCNA creator)"}
I would normalize and transform the data using
`varianceStabilizingTransformation` from `DESeq2`, then cluster samples
and/or calculate latent factors such as principal components or
Surrogate Variables (R package `sva`) and see whether sequencing depth
has a large effect (whether it correlates with the leading latent
factors). If it does not, I would not worry about it; if it does, you
can try to adjust the data but it's questionable whether linear models
are a suitable way to model the effect of sequencing depth that
remains after normalization.
:::

## Normalization

According to [post](https://www.biostars.org/p/459013/#459016) `varianceStabilizingTransformation` not `vst` is used. Design is provided here only because it is required in `dds` object. Here full data are explored but eventually it will be analyzed separately by tissue. Also, it makes no sense using `blind=FALSE` here.

```{r normalize, cache=TRUE}
vsd=varianceStabilizingTransformation(dds.f, blind=TRUE)
```

# Explorarive data analysis

```{r packages-visual, cache=TRUE}
#| output: false
#| message: false
library(pheatmap)
library(RColorBrewer)
library(viridis)
library(PoiClaClu)
library(DESeq2)
library(ggplot2)
```

## Heatmaps

```{r dists, cache=TRUE}
#| output: false
#| message: false
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix( sampleDists )
```

```{r heat-basic, fig.width=10, fig.height=10, fig.dpi=600}
#| eval: false
rownames(sampleDistMatrix) <- paste( vsd$tissue, vsd$line, vsd$time, vsd$day, vsd$rep, sep = " - " )
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
png("heat.png", width=50, height=50, units="cm", res=150)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
dev.off()
```

```{r heat-basic2, fig.width=10, fig.height=10, fig.dpi=600}
#| eval: false
rownames(sampleDistMatrix) <- colnames(vsd)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
png("heat-id.png", width=50, height=50, units="cm", res=150)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
dev.off()
```

The two above plots validate `coldata` correctness.
As in cold + regrowth experiment, samples are clustered primarily by tissue type and then by inbred line.
Interestingly in leaf the 3rd level is time of the day (`time`) what is not visible in SAM samples. This is visualised in `heat-line-time` chunk.

```{r heat-line-time, fig.width=10, fig.height=10, fig.dpi=600}
# names
rownames(sampleDistMatrix) <- colnames(vsd)
colnames(sampleDistMatrix) <- colnames(vsd)

# annotation
annotation_col <- data.frame(
  line = vsd$line,
  time = vsd$time
)
rownames(annotation_col) <- colnames(vsd)

# colors
ann_colors <- list(
  line = setNames(brewer.pal(length(unique(vsd$line)), "Set2"), unique(vsd$line)),
  time = setNames(brewer.pal(length(unique(vsd$time)), "Pastel1"), unique(vsd$time))
)

colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

# heatmap
#png("heat-line-powt.png", width=50, height=50, units="cm", res=150)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         annotation_colors = ann_colors)
#dev.off()
```

```{r heat-day, fig.width=10, fig.height=10, fig.dpi=600}
# names
rownames(sampleDistMatrix) <- colnames(vsd)
colnames(sampleDistMatrix) <- colnames(vsd)

# annotation
day <- data.frame(day = vsd$day)
rownames(day) <- colnames(vsd)  # match names

# colors
day_levels <- unique(day$day)
day_colors <- setNames(RColorBrewer::brewer.pal(length(day_levels), "Set2"), day_levels)
ann_colors <- list(day = day_colors)

# heatmap
#png("heat-day.png", width=50, height=50, units="cm", res=150)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         annotation_col = day,
         annotation_row = day,
         annotation_colors = ann_colors)
#dev.off()
```

```{r heat-rep, fig.width=10, fig.height=10, fig.dpi=600}
# names
rownames(sampleDistMatrix) <- colnames(vsd)
colnames(sampleDistMatrix) <- colnames(vsd)

# annotation
rep <- data.frame(rep = vsd$rep)
rownames(rep) <- colnames(vsd)  # match names

# colors
rep_levels <- unique(rep$rep)
rep_colors <- setNames(RColorBrewer::brewer.pal(length(rep_levels), "Set2"), rep_levels)
ann_colors <- list(rep = rep_colors)

# heatmap
#png("heat-rep.png", width=50, height=50, units="cm", res=150)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         annotation_col = rep,
         annotation_row = rep,
         annotation_colors = ann_colors)
#dev.off()
```

```{r heat-Mreads-rep, fig.width=10, fig.height=10, fig.dpi=600}
rownames(sampleDistMatrix) <- colnames(vsd)
colnames(sampleDistMatrix) <- colnames(vsd)

# Column with "millions of reads"
reads <- vsd$Mreads

# Binning to 7 equal bins
reads_bin <- cut(reads, breaks = 4, include.lowest = TRUE)

# Column annotation
annotation_col <- data.frame(reads_bin = reads_bin,
                             rep = vsd$rep)
rownames(annotation_col) <- colnames(vsd)

# Nice continous palette with contrast + good for people with colour vision defects

reads_colors <- setNames(
  viridis::viridis(4, option = "D"),  # options: "D", "C", "A", etc.
  levels(reads_bin)
)

rep_levels <- unique(rep$rep)
rep_colors <- setNames(RColorBrewer::brewer.pal(length(rep_levels), "Set2"), rep_levels)
# For annotation_colors
annotation_colors <- list(reads_bin = reads_colors,
                          rep = rep_colors)

# plot
#png("heat-reads-a.png", width=50, height=50, units="cm", res=150)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         annotation_colors = annotation_colors)
#dev.off()
```

Neither `day`, `rep` nor `Mreads` are separating samples. Also, `rep` and are not confounded with `Mreads`.

### Heatmaps with Poisson distance

::: {.callout-note}
Heatmaps generated with Poisson distance does not show nothing new - they'll be hidden.
:::

```{r dists-poi, cache=TRUE}
#| output: false
#| message: false
#| eval: false
#| include: false
poisd <- PoissonDistance(t(counts(dds.f)))
samplePoisDistMatrix <- as.matrix( poisd$dd )
```

```{r poi-basic, fig.width=10, fig.height=10, fig.dpi=600}
#| eval: false
#| include: false
rownames(samplePoisDistMatrix) <- paste( vsd$tissue, vsd$line, vsd$time, vsd$day, vsd$rep, sep = " - " )
colnames(samplePoisDistMatrix) <- NULL
#png("poi-basic.png", width=50, height=50, units="cm", res=150)
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = poisd$dd,
         clustering_distance_cols = poisd$dd,
         col = colors)
#dev.off()
```

```{r poi-basic2, fig.width=10, fig.height=10, fig.dpi=600}
#| eval: false
#| include: false
rownames(samplePoisDistMatrix) <- colnames(vsd)
colnames(samplePoisDistMatrix) <- NULL
#png("poi-basic2.png", width=50, height=50, units="cm", res=150)
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = poisd$dd,
         clustering_distance_cols = poisd$dd,
         col = colors)
#dev.off()
```

```{r poi-Mreads-rep, fig.width=10, fig.height=10, fig.dpi=600}
#| eval: false
#| include: false
rownames(samplePoisDistMatrix) <- colnames(vsd)
colnames(samplePoisDistMatrix) <- colnames(vsd)

# Column with "millions of reads"
reads <- vsd$Mreads

# Binning to 7 equal bins
reads_bin <- cut(reads, breaks = 4, include.lowest = TRUE)

# Column annotation
annotation_col <- data.frame(reads_bin = reads_bin,
                             rep = vsd$rep)
rownames(annotation_col) <- colnames(vsd)

# Nice continous palette with contrast + good for people with colour vision defects

reads_colors <- setNames(
  viridis::viridis(4, option = "D"),  # options: "D", "C", "A", etc.
  levels(reads_bin)
)

rep_levels <- unique(rep$rep)
rep_colors <- setNames(RColorBrewer::brewer.pal(length(rep_levels), "Set2"), rep_levels)
# For annotation_colors
annotation_colors <- list(reads_bin = reads_colors,
                          rep = rep_colors)

# plot
#png("heat-reads-a.png", width=50, height=50, units="cm", res=150)
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         annotation_col = annotation_col,
         annotation_row = annotation_col,
         annotation_colors = annotation_colors)
#dev.off()
```

```{r poi-line, fig.width=10, fig.height=10, fig.dpi=600}
#| eval: false
#| include: false
# names
rownames(samplePoisDistMatrix) <- colnames(vsd)
colnames(samplePoisDistMatrix) <- colnames(vsd)

# annotation
line <- data.frame(line = vsd$line)
rownames(line) <- colnames(vsd)  # match names

# colors
line_levels <- unique(line$line)
line_colors <- setNames(RColorBrewer::brewer.pal(length(line_levels), "Set2"), line_levels)
ann_colors <- list(line = line_colors)

#png("poi-line.png", width=50, height=50, units="cm", res=150)
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         annotation_col = line,
         annotation_row = line,
         annotation_colors = ann_colors)
#dev.off()
```

## PCA plots

Dataset is partitioned primarily by tissue
```{r pca-tissue, cache=TRUE}
plotPCA(vsd, intgroup = c("tissue"), ntop=33000) +
  ggtitle("PCA – color by tissue") +
  theme(plot.title = element_text(hjust = 0.5))
```
Separation by line is also visible in PC2
```{r pca-line, cache=TRUE}
plotPCA(vsd, intgroup = c("line"), ntop=33000) +
  ggtitle("PCA – color by inbred line") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r pca-pcs23, cache=TRUE}
# Code to see the separation of samples in PC2 and PC3
# Little variance
plotPCA(vsd, intgroup = c("line"), ntop=33000, pcs=2:3) +
  ggtitle("PCA – color by line") +
  theme(plot.title = element_text(hjust = 0.5))
```

Slight (low variance) separation by `line` is also visible in PC2 and PC3 but is negligible and `tissue` effect is dominating.
So tissues should be analyzed separately.

# Splitting of data by tissue

```{r split-data, cache=TRUE}
# https://bioconductor.org/packages/devel/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html#subsetting
vsd.leaf=vsd[, vsd$tissue == 'leaf']
vsd.sam=vsd[, vsd$tissue == 'sam']

# check
dim(vsd.leaf)
dim(vsd.sam)
```

Datasets are separated by tissue
```{r pca-tissues, cache=TRUE}
#| layout-ncol: 2
plotPCA(vsd.leaf, intgroup = c("line"), ntop=33000) +
  ggtitle("PCA for leaf – color by inbred line") +
  theme(plot.title = element_text(hjust = 0.5))
plotPCA(vsd.sam, intgroup = c("line"), ntop=33000) +
  ggtitle("PCA for sam – color by inbred line") +
  theme(plot.title = element_text(hjust = 0.5))
```

Neither of remaining factors (time, day, rep, batch, million reads) are separating samples so it is not shown.

It is noteworthy, that variance of PC3 is not very smaller than that of PC2. PC4 explains only 5% of variance in SAM and 9% in leaf.
S311 separates somehow from others.

```{r pca23-line, cache=TRUE}
#| layout-ncol: 2
plotPCA(vsd.leaf, intgroup = c("line"), ntop=33000, pcs=2:3) +
  ggtitle("PCA for leaf – color by inbred line") +
  theme(plot.title = element_text(hjust = 0.5))
plotPCA(vsd.sam, intgroup = c("line"), ntop=33000, pcs=2:3) +
  ggtitle("PCA for sam – color by inbred line") +
  theme(plot.title = element_text(hjust = 0.5))
```
