---
title: "Pairwise comparisons for each line pairs across each timepoint, SAM, control."
format: html
date: "2026-03-24"
bibliography: refs-stat-anal.bib
execute:
  eval: false
  fig-format: png
  fig.dpi: 300
---

Built with R version `r getRversion()`

## Rationale

Using only subset of data for sam and design matrix to extract interesting comparisons.

The experiment entailed three days and have three replications.

d1 - d2 - d3

In each day we have three samples: dawn, day, and dusk.

d1[dawn, day, dusk] - d2[dawn, day, dusk] - d3[dawn, day, dusk]

And this will be recoded as 1 - 9.

In all there are 9 time points and 9 pairwise comparisons. This scheme is repeated for four maize inbred lines.

So in all there are 9 (timepoints) x 6 (line pairs) = 54 comparisons

`DESeq2` package will be used [@loveModeratedEstimationFold2014].
The analysis is based on [RNA-seq workflow at the gene level](https://bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html) and [Analyzing RNA-seq data with DESeq2](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#differential-expression-analysis)

## Environment cleaning and set-up

```{r setup, include = FALSE}
# load required packages
library(DESeq2)
library(GenomicRanges)
library(BiocParallel)
library(limma)
library(dplyr)
library(tibble)
```

```{r install packages}
#| eval: false
#| output: false
#| message: false
# Install required packages if not done previously
## From Bioconductor
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GenomicRanges")
BiocManager::install("DESeq2")
BiocManager::install("genefilter")
BiocManager::install("apeglm")

## From CRAN
install.packages("pheatmap")
install.packages("RColorBrewer")
install.packages("viridis")
install.packages("PoiClaClu")
install.packages("glmpca")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("ashr")
```

```{r}
#| eval: false
#| output: false

# Clean working directory
rm(list=ls())
```

## Data preprocessing

### Load raw count data

Full dataset for control

```{r read data, cache=TRUE}
load("../../../input-data/rdata/counts_and_metadata.RDa")

# previously prepared range data
load("../../../../project1-stat/input-data/rdata/ranges.RDa")

# Check if files are in concert
all(rownames(coldata) %in% colnames(cts))
all(rownames(coldata) == colnames(cts))
```

### Subset sam data

Subsetting is done at the level of raw data (counts and coldata).
```{r subset sam data}
# subset coldata to contain only sam data
coldata.sam=subset(coldata, tissue=="sam")

# check
dim(coldata.sam)

# add variable describing chronology of samples
coldata.sam$timepoint = as.integer(as.numeric(coldata.sam$time)) + ( as.integer(as.numeric(coldata.sam$day)) - 1 ) * 3
# make 'timepoint' a factor
coldata.sam$timepoint = as.factor(coldata.sam$timepoint)

# Adding grouping variable to easily analyse all comparisons needed
coldata.sam$group <- factor(paste0(coldata.sam$line, "_", coldata.sam$timepoint))

# vector of sample names to keep - needed to select columns from count file
keep.sam=rownames(coldata.sam)

# check - only samples named "s..." should remain
keep.sam
length(keep.sam)

# subsetting counts
cts.sam=subset(cts, select=keep.sam)

# check
dim(cts.sam)
colnames(cts.sam)

# Check if counts and coldata are in concert
all(rownames(coldata.sam) %in% colnames(cts.sam))
all(rownames(coldata.sam) == colnames(cts.sam))

keep = c("coldata.sam", "cts.sam", "gzakresy")

rm(list = setdiff(ls(), keep))
```

### Create DESeq2 object for sam only

```{r create DESeq2 object, cache=TRUE}
library("DESeq2")
#| output: false
#| message: false

# Design according to:
# https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#time-course-experiments

dds=DESeqDataSetFromMatrix(countData=cts.sam, colData=coldata.sam, rowRanges=gzakresy, design= ~ 0 + group)

```

### Pre-filtering

```{r pre-filtering, cache=TRUE}
# at least 3 samples with a count of 10 or higher
keep <- rowSums(counts(dds) >= 10) >= 3
dds.f <- dds[keep,]

# check
dim(dds.f)

# what percent of genes remained?
100*(dim(dds.f)[1]/dim(dds)[1])

keep = c("dds.f")

rm(list = setdiff(ls(), keep))
```

## Analysis

### Test


```{r analysis, cache=TRUE}
#| eval: false

# parallelization setup
## otherwise test runs very long

register(MulticoreParam(8))

# analysis using simple design already specified in dds.f
dds.f <- DESeq(dds.f, parallel = TRUE)
```

```{r env cleanup}
# genes from the following object will be used as background in GO enrichment analysis
save(dds.f, file="../rdata-saved/dds.f.RDa")
```

### Selecting interesting comparisons

```{r refresh}
# remove all objects and load only dds.f object (result of DEseq function)
rm(list = ls())

load("../rdata-saved/dds.f.RDa")
```

For logFC shrinkage `ashr` is used [@stephensFalseDiscoveryRates2016] instead of `apeglm` as the latter doesn't support contrasts.

```{r}
# according to post by ATpoint
# https://support.bioconductor.org/p/9139197/#9139238

unique_combn <- combn(c("a554", "s018", "s84", "s311"), 2)

timepoints = c(1 : 9)

allcontrasts = list()

for (timepoint in timepoints){
  contrasts <- lapply(1:ncol(unique_combn), function(x){
  
    c(paste0("group", unique_combn[1,x], "_", timepoint), paste0("group", unique_combn[2,x], "_", timepoint))
  
  })
  
  allcontrasts = append(allcontrasts, contrasts)

}

invisible(
  lapply(allcontrasts, function(contr) {

    contr1 = unlist(contr)
    
    
    res_name <- paste(gsub("group", "", contr1[1]), gsub("group", "", contr1[2]), sep=".")
    
    res=lfcShrink(dds.f, contrast=list(contr1[1], contr1[2]), type="ashr")
    
    assign(res_name, res, envir = .GlobalEnv)
  })
)
```

```{r save-contr}
to_save = ls(pattern = "\\.[sa]")

save(list = to_save, file="../rdata-saved/contr_pair_sam_ctrl.RDa")

rm(list = setdiff(ls(), to_save))
```

```{r sigs}
# extract and save only significant genes
original_names = ls(pattern = "\\.[sa]")

invisible(
  lapply(original_names, function(name) {
    obj <- get(name)
    
    sig_name = paste0(name, "_sam_ctrl_sig")
    
    res <- obj[which(obj$padj < 0.05), ]
    assign(sig_name, res, envir = .GlobalEnv)
    
  })
)

to_save = ls(pattern = "sig$")

# this file will be used for clustering with mfuzz
save(list = to_save, file="../rdata-saved/contr_pair_sam_ctrl_sig005.RDa")

#rm(list = setdiff(ls(), to_save))
```

## Compute and export normalized counts

This was requested by PI.
The computed values are counts with `sizeFactor` applied - allow only between sample comparisons of the same gene.
Other option is TPM - allow between sample (of the same sample group = within cold, within control) within gene and between genes (within sample) comparisons.
TPM may be computed directly by `edgeR`.

```{r normalized}
load("../rdata-saved/dds.f.RDa")

# apply sizeFactors on counts
cts_norm = counts(dds.f, normalized = TRUE)

# make coldata with grouping variable for averaging - used as final label
coldata = colData(dds.f)
# make variable for averaging
coldata$forAverage <- factor(paste0(coldata$line, "_", coldata$timepoint, "_S"))

# Check if files are in concert
all(rownames(coldata) %in% colnames(cts_norm))
all(rownames(coldata) == colnames(cts_norm))

# Steps needed to make expressionSet object
# varMetadata - description of variables
metadata=data.frame(labelDescription=c("pooled sample", "letter", "tissue", "line", "hour of pooling", "sample number", "day", "replication", "condition", "id", "reads [million]", "million of reads binned", "chronological time point", "group", "sizeFactor", "grouping variable for averaging"), row.names=colnames(coldata))

# metadata
phenoData=new("AnnotatedDataFrame", data=as.data.frame(coldata), varMetadata=metadata)
phenoData

## check
# sampleNames(phenoData)
# varMetadata(phenoData)
# head(pData(phenoData))

# make ExpressionSet
dds.eset <- ExpressionSet(assayData=cts_norm, phenoData=phenoData)
```

```{r average}
# average reps
average = t(avereps(t(exprs(dds.eset)), ID = dds.eset$forAverage)) # output is a matrix

# check
dim(average)
dim(dds.eset)

## sort chronologically within line
# get and order colnames

cols <- colnames(average)
prefix <- sub("_.*", "", cols)
num <- as.numeric(sub(".*_([0-9]+).*", "\\1", cols))
ord <- order(prefix, num)
average_srt <- average[, ord]

# save object - it will be joined with SAM. UniProt ID and description will be also added
# save(average_srt, file = '../rdata-saved/leaf_ave_norm_count.RDa')
```

```{r anno}
# make column with gene names
df <- as.data.frame(average_srt) |>
  rownames_to_column(var = "Name")

# load annotation
anno = read.table("../../../../project1-stat/input-data/go-uniprot/NAM_descr_best", header=FALSE, stringsAsFactors = FALSE, sep="\t", quote = "")
colnames(anno) <- c("Name", "description")

# join with DE-data
df_anno <- left_join(df, anno, by = "Name")

# join with UPs
ups = read.table("../../../../project1-stat/input-data/go-uniprot/NAM2UP_all_forGOextraction", header=FALSE, stringsAsFactors = FALSE, sep="\t", quote = "")
colnames(ups) <- c("Name", "UniProt")

average_anno_up <- left_join(df_anno, ups, by = "Name") |>
  relocate(UniProt, .after = Name)

# save to file
write.table(average_anno_up, file = 'SAM_ctrl_ave_count_sizefactor.csv', row.names = FALSE, quote = FALSE, sep='\t')
```


## References
