---
title: "Control codition. Comparison of subsequent time-points to time zero in SAM"
format: html
date: "2026-01-21"
bibliography: refs-stat-anal.bib
execute:
  eval: false
  fig-format: png
  fig.dpi: 300
---

:::{.callout-note}
For html creation purposes I've turned off code execution.
:::

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(root.dir = "/media/mj/ANTIX-LIVE/project1-ctrl-stat/vs-day-one/sam/scripts")
```

Built with R version `r getRversion()`

## Rationale

Using only subset of data for SAM and design matrix to extract interesting comparisons.

The experiment entailed three days and have three replications.

d1 - d2 - d3

In each day we have three samples: dawn, day, and dusk.

d1[dawn, day, dusk] - d2[dawn, day, dusk] - d3[dawn, day, dusk]

And we want to compare:

d2[dawn, day, dusk] *vs* d1[dawn, day, dusk]
d3[dawn, day, dusk] *vs* d1[dawn, day, dusk]

So in all there are 3 time points and we have 6 comparisons.

There are four inbred lines so the same analysis must be done for each.

`DESeq2` package will be used [@loveModeratedEstimationFold2014].
The analysis is based on [RNA-seq workflow at the gene level](https://bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html) and [Analyzing RNA-seq data with DESeq2](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#differential-expression-analysis)

Environment cleaning and set-up
```{r}
#| eval: false
#| output: false
# Clean working directory
rm(list=ls())
```

## Environment set-up

```{r install packages}
#| eval: false
#| output: false
#| message: false
# Install required packages if not done previously
## From Bioconductor
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GenomicRanges")
BiocManager::install("DESeq2")
BiocManager::install("genefilter")
BiocManager::install("apeglm")

## From CRAN
install.packages("pheatmap")
install.packages("RColorBrewer")
install.packages("viridis")
install.packages("PoiClaClu")
install.packages("glmpca")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("ashr")
```

## Data preprocessing

### Load raw count data

Full dataset from control

```{r read data, cache=TRUE}
load("../../../input-data/rdata/counts_and_metadata.RDa")

# previously prepared range data
library("GenomicRanges")
load("../../../../project1-stat/input-data/rdata/ranges.RDa")

# Check if files are in concert
all(rownames(coldata) %in% colnames(cts))
all(rownames(coldata) == colnames(cts))
```

### Subset SAM data

Subsetting is done at the level of raw data (counts and coldata).
```{r subset SAM data}
# subset coldata to contain only sam data
coldata.sam=subset(coldata, tissue=="sam")

# check
dim(coldata.sam)

# vector of sample names to keep - needed to select columns from count file
keep.sam=rownames(coldata.sam)

# check - only samples named "s..." should remain
keep.sam
length(keep.sam)

# subsetting counts
cts.sam=subset(cts, select=keep.sam)

# check
dim(cts.sam)
colnames(cts.sam)

# Check if counts and coldata are in concert
all(rownames(coldata.sam) %in% colnames(cts.sam))
all(rownames(coldata.sam) == colnames(cts.sam))
```

### Create DESeq2 object for SAM only

```{r create DESeq2 object, cache=TRUE}
library("DESeq2")
#| output: false
#| message: false

# Adding grouping variable to easily analyse all comparisons needed
coldata.sam$group <- factor(paste0(coldata.sam$line, "_d", coldata.sam$day, "_t", coldata.sam$time))

# simple design with group
# according to 
# https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#interactions

dds=DESeqDataSetFromMatrix(countData=cts.sam, colData=coldata.sam, rowRanges=gzakresy, design= ~ 0 + group)

```

### Pre-filtering

```{r pre-filtering, cache=TRUE}
# at least 3 samples with a count of 10 or higher
keep <- rowSums(counts(dds) >= 10) >= 3
dds.f <- dds[keep,]

# check
dim(dds.f)

# what percent of genes remained?
100*(dim(dds.f)[1]/dim(dds)[1])
```

## Analysis

### Test


```{r analysis, cache=TRUE}
#| eval: false

# parallelization setup
## otherwise test runs very long

library("BiocParallel")
register(MulticoreParam(8))

# analysis using simple design already specified in dds.f
dds.f <- DESeq(dds.f, parallel = TRUE)
```

### Environment cleanup

```{r env cleanup}
# save workspace
save.image("../rdata-saved/deseq_cold_SAM.RDa")

# remove everything except for the DESeq2 object with results
rm(list= ls()[!(ls() %in% c('dds.f'))])

# genes from the following object will be used as background in GO enrichment analysis
save(dds.f, file="../rdata-saved/dds.f.RDa")
```

### Selecting interesting comparisons

For logFC shrinkage `ashr` is used [@stephensFalseDiscoveryRates2016] instead of `apeglm` as the latter doesn't support contrasts.

```{r contrasts}
#| eval: false
# using log fold change shrinkage (substitutes results function)

# vector of lines and times

lines = c("a554", "s018", "s84", "s311")

times = c("t1", "t2", "t3")

for (line in lines){
  for (time in times){
  
    # Names from resultsNames
    
    trt_d2 <- paste0("group", line, "_d2_", time)
    trt_d3 <- paste0("group", line, "_d3_", time)
    
    ref <- paste0("group", line, "_d1_", time)
    
    # Names for result objects
    
    res_d2 <- paste0(line, "_d2d1_", time)
    res_d3 <- paste0(line, "_d3d1_", time)
    
    ## day 2 vs day 1
    
    res=lfcShrink(dds.f, contrast=list(trt_d2, ref), type="ashr")
    
    # Check
    summary(res)
    
    # Save object to .GlobalEnv
    assign(res_d2, res, envir = .GlobalEnv)
    
    ## day 3 vs day 1
    
    res=lfcShrink(dds.f, contrast=list(trt_d3, ref), type="ashr")
    
    # Check
    summary(res)
    
    # Save object to .GlobalEnv
    assign(res_d3, res, envir = .GlobalEnv)
    
  }
}
```
### Extracting significant results

### Subset data to upregulated and downregulated genes

```{r pl min}
original_names <- grep("^[as].*t[1-3]$", ls(), value = TRUE)

invisible(
  lapply(original_names, function(name) {
    obj <- get(name)

    pl_res_name <- paste0(name, "_sam_pl")
    min_res_name <- paste0(name, "_sam_min")
    
    sig_name = paste0(name, "_sam_sig")
    
    # all
    res <- obj[which(obj$padj < 0.05), ]
    assign(sig_name, res, envir = .GlobalEnv)
    
    # plus
    res <- obj[which(obj$padj < 0.05 & obj$log2FoldChange > 0), ]
    assign(pl_res_name, res, envir = .GlobalEnv)

    # minus
    res <- obj[which(obj$padj < 0.05 & obj$log2FoldChange < 0), ]
    assign(min_res_name, res, envir = .GlobalEnv)

  })
)
```

### Save results to object for GO analysis

```{r save results}
# # select pattern
# to_save <- ls(pattern = "sam_(pl|min)$")
# # check
# to_save
# 
# # saving
# save(list = to_save, file="../rdata-saved/sigs_SAM1.RDa")

# # also save unsplitted results
# to_save = ls(pattern = "t[1-3]$")
#
# to_save
#
# # saving
# save(list = to_save, file="../rdata-saved/sigs-plmin.RDa")

# Only unsplitted significant genes are needed further

# select pattern
to_save <- ls(pattern = "_sig$")
# check
to_save

# saving
save(list = to_save, file="../rdata-saved/sigs_all_SAM.RDa")
```

<!-- Following code can be customized to select genes with desired fold change for clustering. -->

<!-- ```{r FC0.5} -->
<!-- # list of gene with minimal FC of 0.5 in either direction -->
<!-- # to use in Mfuzz clustering -->

<!-- load("../rdata-saved/sigs_all_SAM1.RDa") -->
<!-- original_names <- grep("^[as].*sig$", ls(), value = TRUE) -->

<!-- invisible( -->
<!--   lapply(original_names, function(name) { -->
<!--     obj <- get(name) -->

<!--     sig_name = paste0(name, "_fc05") -->

<!--     res <- obj[which(obj$padj < 0.05 & abs(obj$log2FoldChange) > 0.5), ] -->
<!--     assign(sig_name, res, envir = .GlobalEnv) -->
<!--   }) -->
<!-- ) -->

<!-- # select pattern -->
<!-- to_save <- ls(pattern = "_sam_sig_fc05$") -->
<!-- # check -->
<!-- to_save -->

<!-- # saving -->
<!-- save(list = to_save, file="../rdata-saved/sigs_fc05.RDa") -->
<!-- ``` -->

## References
