---
title: "Prepare network files for analysis in Cytoscape, control"
format: html
date: "2026-02-09"
bibliography: ../../../refs-stat-anal.bib
execute:
  fig-format: png
  fig.dpi: 300
---

The aim of this code is to prepare files (clustering results and String network data [@szklarczykSTRINGDatabase20232023]) for Cytoscape analysis and visualization [@clineIntegrationBiologicalNetworks2007]. Separately for each cluster.
Using 0.7 membership cutoff.
This script uses GO-annotation and associated files from UniProt, prepared in `../../../input-data/go-uniprot/00-test-filter.qmd` and `../../../input-data/go-uniprot/00-GO-and-mapping-UniProt.qmd`.
Here option 1) using NAM - GO is realized

Built with R version `r getRversion()`

## R Preliminaries

Environment cleaning and set-up

```{r packages}
#| eval: false
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Mfuzz")
BiocManager::install("marray")
BiocManager::install("limma")
```

```{r}
#| output: false

# Clean working directory
rm(list=ls())
```

## Load data

`c10_m1.25` contains data for clusters.


```{r load}
load("../rdata-saved/mfuzz_c8_m1.15_samALLsig.RDa")

# gene description
anno = read.table("../../../../project1-stat/input-data/go-uniprot/NAM_descr_best", header=FALSE, stringsAsFactors = FALSE, sep="\t", quote = "")
colnames(anno) <- c("gene", "description")

```

## Prepare and export cluster data

```{bash net-dir}
rm -r ../results/network/
mkdir ../results/network/
```

```{r membership-table}
## All genes in clusters with membership values
## 1.  Sprawdź, co w nim jest
# str(c8_m1.15)

## 2.  Pobierz wektor cluster (nazwany) i macierz membership
cl_vec   <- c8_m1.15$cluster               # wektor długości 11936, nazwy = geny
mem_mat  <- c8_m1.15$membership            # macierz 11936 × 10

## 3.  Upewnij się, że nazwy wierszy obu obiektów są zgodne
identical(names(cl_vec), rownames(mem_mat))

## 4.  Stwórz data‑frame z dwoma zestawami kolumn
df <- data.frame(
  cluster = cl_vec,                     # pierwsza kolumna – numer klas
  mem_mat,                              # kolejne 10 kolumn – membership
  check.names = FALSE                   # pozwala na nazwę "1", "2", …
)

## 5.  (Opcjonalnie) nadaj sensowne nazwy kolumnom membership
colnames(df)[2:ncol(df)] <- paste0("mem_", seq_len(ncol(mem_mat)))

# make column from rownames

library(dplyr)

df2 <- df |>
  mutate(gene = rownames(df)) |>   # nowa kolumna "gene"
  select(gene, everything())             # opcjonalnie przestaw kolejność

# select only rows where any of the memberships is >= 0.7
c8_m1.15_mem07 <- df2 |>
  filter(
    if_any(starts_with("mem_"),
           ~ . >= 0.7)              # warunek na każdą z nich
  )


# Extract max value from membership columns and retain it in new column
# 1 Wybieramy kolumny zaczynające się od "mem_"
mem_cols <- grep("^mem_", names(c8_m1.15_mem07), value = TRUE)

# 2 Obliczamy maksimum w każdym wierszu (ignorujemy NA)
max_mem <- apply(c8_m1.15_mem07[mem_cols], 1, max, na.rm = TRUE)

# 3 Tworzymy nową ramkę z wybranymi kolumnami + maksymalną wartością
c8_m1.15_maxmem <- data.frame(
  gene        = c8_m1.15_mem07$gene,
  cluster     = c8_m1.15_mem07$cluster,
  max_mem     = max_mem,
  stringsAsFactors = FALSE
)

# Table will be joined with String network data by gene name so it must be sorted

c8_m1.15_sort <- c8_m1.15_maxmem[order(c8_m1.15_maxmem$gene), ]

write.table(c8_m1.15_sort, file="../results/network/c8_m1.15_sort", sep="\t", quote=FALSE, row.names=FALSE)

```

## Add UniProt-NAM mapping, prepared in `00-GO-and-mapping-UniProt.qmd`

Preferentially UPs in STRING were selected already.


```{bash clean-network}
##### This step must be done only once! #####
#
# removing unnecessary prefix from IDs in whole network
## This step uses source file stored outside project directory tree and temporarily use home dir
#zcat /media/mj/ANTIX-LIVE/siec-ala-string/4577.protein.links.full.v12.0.txt.gz | sed 's/4577\.//g' > ~/full-network
#pigz --best ~/full-network
#mv ~/full-network.gz ../../../input-data/networks-up/
#
#############################################
```

```{bash UPs-NAM}
##### This step must be done only once! #####
# 
# cd ../../../input-data/networks-up/
# 
# # change order of columns
# awk -F'\t' -v OFS='\t' '{print $2,$1}' ../go-uniprot/NAM2UP_all_forGOextraction > best-nam-string-ids
# 
# # prepare dictionary for sed
# awk -F'\t' '{print "s/"$2"/"$1"/g"}' best-nam-string-ids > nam-up-dict
#############################################
```

## Join cluster data with UP-NAM mapping

```{bash join-cluster-nam}
##### This step must be done (STEP BY STEP) only once for a given 'c' and 'm' combination! #####

cd ../results/network/
  
# join and sort by UP
join -1 2 -2 1 -t $'\t' ../../../../../project1-stat/input-data/networks-up/best-nam-string-ids c8_m1.15_sort | sort -k2,2 -t $'\t'> c8_UP_srt

# checks - NAMs are unique but UP no
wc -l c8_UP_srt
# 10365
cut -f1 -d $'\t' c8_UP_srt | sort -u | wc -l
# 10365
cut -f2 -d $'\t' c8_UP_srt | sort -u | wc -l
# 10332 # some UPs map to multiple NAMs

# Check
cut -f2 -d $'\t' c8_UP_srt | sort | uniq -d > x
grep -Fwf x c8_UP_srt
# Zm00001eb082960 A0A804MG65      7       0.967184997857515
# Zm00001eb091870 A0A804MG65      7       0.898093437301819
# Zm00001eb257230 A0A804PNG1      1       0.700556559623972
# Zm00001eb419090 A0A804PNG1      1       0.822396958267924
# Zm00001eb437710 A0A804RN54      4       0.847835709698504
# Zm00001eb440080 A0A804RN54      1       0.968879603861032
# Zm00001eb025010 A0A804UH07      2       0.999757586330503
# Zm00001eb356570 A0A804UH07      8       0.999994754780813
# Zm00001eb015200 B4FBW7  1       0.87148043079548
# Zm00001eb111740 B4FBW7  7       0.9969238246223
# Zm00001eb018990 B4FCH2  4       0.925926712105715
# Zm00001eb269650 B4FCH2  7       0.839219775415979
# Zm00001eb378250 B4FCH2  7       0.986814381656861
# Zm00001eb082930 B4FFZ6  7       0.993420762255247
# Zm00001eb091900 B4FFZ6  1       0.926396726675927
# Zm00001eb056890 B4FH55  7       0.953376007306081
# Zm00001eb215110 B4FH55  7       0.99486474125787
# Zm00001eb287110 B4FLA3  1       0.94540012433219
# Zm00001eb350780 B4FLA3  7       0.908472524997436
# Zm00001eb364660 B4FNC9  6       0.991760256365602
# Zm00001eb364680 B4FNC9  6       0.997792028218664
# Zm00001eb117140 B6SGR7  6       0.999077516162661
# Zm00001eb300890 B6SGR7  6       0.999420898021692
# Zm00001eb285220 B6SKA7  1       0.973921877194596
# Zm00001eb285230 B6SKA7  1       0.997310257484945
# Zm00001eb072850 P62787  7       0.999980744500855
# Zm00001eb099930 P62787  7       0.998634061996221
# Zm00001eb123370 P62787  7       0.997837871204173
# Zm00001eb187430 P62787  1       0.872449216250567
# Zm00001eb313720 P62787  7       0.998963710384959
# Zm00001eb313730 P62787  7       0.999771625198861
# Zm00001eb321260 P62787  7       0.996883166690617
# Zm00001eb336310 P62787  7       0.984193588408265
# Zm00001eb351700 P62787  7       0.999974916660991
# Zm00001eb428040 P62787  7       0.999941264685691
# Zm00001eb083380 P69246  7       0.999955089835662
# Zm00001eb123830 P69246  7       0.865627664477731
# Zm00001eb137500 P69246  2       0.998257086918131
# Zm00001eb181740 P69246  7       0.999972079386359
# Zm00001eb271240 P69246  7       0.999933995726167
# Zm00001eb376150 P69246  7       0.999752945913957
# Zm00001eb422190 P69246  7       0.999404417460019
# Zm00001eb422440 P69246  7       0.996888657556251
# Zm00001eb205660 Q08069  2       0.995484319127526
# Zm00001eb205670 Q08069  1       0.997292900586657
# Zm00001eb205710 Q08069  7       0.948880138854961
# Zm00001eb202910 Q9ZQX9  1       0.980370957221147
# Zm00001eb214990 Q9ZQX9  7       0.982051633384974
# Zm00001eb241090 Q9ZQX9  7       0.789792721541177

# A0A804RN54, A0A804UH07, B4FBW7, B4FCH2, B4FFZ6, B4FLA3, P62787, P69246, Q08069, Q9ZQX9 duplicated NAMs are in different clusters

# exclude them
echo -e 'A0A804RN54\nA0A804UH07\nB4FBW7\nB4FCH2\nB4FFZ6\nB4FLA3\nP62787\nP69246\nQ08069\nQ9ZQX9' > x

grep -Fwvf x c8_UP_srt > x2
mv x2 c8_UP_srt
```

## Add network data

```{bash join-net-cluster}
cd ../results/network/
  
## extract data for first interactors from network file
  
# cut UPs column, this was done previously BUT this time it won't contain duplicates in different clusters
cut -f2 -d $'\t' c8_UP_srt | sort -u > c8_ids_up

# extract interactions for selected UPs from full network
zgrep -Fwf c8_ids_up ../../../../../project1-stat/input-data/networks-up/full-network.gz > c8_UP_net_selected


## Add NAMs and description ##
# for UP1

# translate spaces to tabs in network data
tr ' ' '\t' < c8_UP_net_selected > ~/x
# sort by UP1
sort -k1,1 -t $'\t' ~/x > ~/xUP1

# join, keeping all network data, there are multiple UPs for some NAMs so number of rows in output is greater than in xUP1
join -1 1 -2 2 -t $'\t' -a1 -o auto ~/xUP1 ../../../../../project1-stat/input-data/go-uniprot/NAM-UP-anno > ~/x2
# 1=UP1, 2=UP2, 17=NAM1, 18=descr1

# for UP2
# sort by UP2
sort -k2,2 -t $'\t' ~/x2 > ~/xUP2
# join - using HOME directory
join -j 2 -t $'\t' -a1 -o auto ~/xUP2 ../../../../../project1-stat/input-data/go-uniprot/NAM-UP-anno > ~/xUPsNAMs
# 1,19,20 - UP2, 2,17,18 - UP1

## Add cluster data ##

# UP1
# check sorting, using original results from R (without UPs)
sort -k1,1 -t $'\t' -c c8_m1.15_sort
# ok

# sort cluster data by NAM1
sort -k17,17 -t $'\t' ~/xUPsNAMs > ~/xNAM1

# join
join -1 17 -2 1 -t $'\t' -a1 -o auto ~/xNAM1 c8_m1.15_sort > ~/x2NAM1
# 1,21,22 - NAM1, 

# UP2
# sort
sort -k19,19 -t $'\t' ~/x2NAM1 > ~/xNAM2

# join
join -1 19 -2 1 -t $'\t' -a1 -o auto ~/xNAM2 c8_m1.15_sort > ~/x2NAM1NAM2
# 1,3,20,23,24 NAM2, 2,4,19,21,22 NAM1

# select rows with NAM1 not empty - this is in the clusters
awk -F'\t' '$21!=""' ~/x2NAM1NAM2 > ~/x3NAM # było $2

# tidy-up
awk -F'\t' -v OFS='\t' '{print $4,$2,$19,$21,$22,$3,$1,$20,$23,$24,$5,$6,$7,$8,$9,$10,$11,$12,$13,$14,$15,$16,$17,$18}' ~/x3NAM > ~/x4final
```

```{bash checks}
cd ../results/network/
## Checks
# NAM - UP mapping
# cut NAMs and UPs for both interacting proteins
cut -f1,2 -d $'\t' ~/x4final > ~/x1
cut -f6,7 -d $'\t' ~/x4final > ~/x2
# concatenate
cat ~/x1 ~/x2 | sort -u > ~/x3
# translate NAMs to UPs
sed -f ../../../../../project1-stat/input-data/networks-up/nam-up-dict ~/x3 > ~/x
# select rows where UP2 are missing
awk -v FS='\t' -v OFS='\t' '$1!=$2 {print $2}' ~/x | sort -u
# should be empty

# NAM - description
# cut corresponding columns
cut -f2,3 -d $'\t' ~/x4final | sort -u > ~/x1
cut -f7,8 -d $'\t' ~/x4final | sort -u > ~/x2
# concatenate
cat ~/x1 ~/x2 | sort -u > ~/x3
# compare to file with description
join -j1 -t $'\t' ../../../../../project1-stat/input-data/go-uniprot/NAM_descr_best ~/x3 > ~/x4
# select rows with different data in description
awk -v FS='\t' -v OFS='\t' '$2!=$3' ~/x4 | sort -u
# should be empty

## Check NAM - cluster assignment
# extract respecive columns
cut -f2,4,5 -d $'\t' ~/x4final | sort -u > ~/x1
cut -f7,9,10 -d $'\t' ~/x4final | sort -u > ~/x2
# concatenate
cat ~/x1 ~/x2 | sort -u > ~/x3
# does data present in both files match?
join -j1 -t $'\t' c8_m1.15_sort ~/x3 > ~/x4
# cut respective columns and compare with diff
cut -f2,3 -d $'\t' ~/x4 > ~/x41
cut -f4,5 -d $'\t' ~/x4 > ~/x42
diff -s ~/x41 ~/x42
# Pliki /home/mj/x41 i /home/mj/x42 są identyczne

# Check NAMs not in cluster
# extract NAMs without cluster from final file
join -j1 -t $'\t' -a1 -a2 -o auto c8_m1.15_sort ~/x3 | awk -v FS='\t' -v OFS='\t' '$2=="" && $4==""' > ~/x5
# grep them from network assignment file
grep -Fwf ~/x5 c8_m1.15_sort > ~/x2
# check number of rows, should be 0
wc -l ~/x2
# 0 /home/mj/x2

## Check UP - UP assignment
# extract from final file
cut -f1,6 -d $'\t' ~/x4final | tr -d '\t' | sort -u > ~/x
# extract info from full network
zcat ../../../../../project1-stat/input-data/networks-up/full-network.gz | cut -f1,2 -d' ' | tr -d ' ' | tail -n +2 | sort -u > ~/x2
# grep data from final file from full network - all should be there
grep -Fwf ~/x ~/x2 > ~/x3
# compare
diff -s  ~/x3 ~/x
# Pliki /home/mj/x3 i /home/mj/x są identyczne

# Cleanup

mv ~/x4final c8_net_allcl
rm ~/x*

##  TU SKOŃCZYŁEM ##

# Manual checks for random entries
shuf -n3 c8_net_allcl

### Examples
## input-data/go-uniprot dir
# NAM - UP assignment
grep Zm00001eb314790 BEST_UPs
grep A0A1D6LKH6 BEST_UPs

# NAM and UP description
grep Zm00001eb269840 NAM_descr_best
grep A0A1D6LKH6 UP_descr_best

## Cluster assignment
# sam/results/network dir
grep Zm00001eb269840 c8_m1.15_sort

```

## Split by cluster

```{bash split}
cd ../results/network/

# using 400 score cutoff
awk -v FS='\t' -v OFS='\t' '$24>=400 { print > "cluster_"$4"_400.txt" }' c8_net_allcl

# add header

for i in {1..8}; do
  echo -e "uniprot1\tensembl1\tdesc1\tcluster1\tmember1\tuniprot2\tensembl2\tdesc2\tcluster2\tmember2\tneighborhood\tneighborhood_transferred\tfusion\tcooccurence\thomology\tcoexpression\tcoexpression_transferred\texperiments\texperiments_transferred\tdatabase\tdatabase_transferred\ttextmining\ttextmining_transferred\tcombined_score" | cat - cluster_${i}_400.txt > temp && mv temp cluster_${i}_400.txt ;
done

pigz --best c8_net_allcl
```

:::{.callout-important}
The rest of the code is probably superfluous. Further extended set of GO relations is used.
Transition to [10-NAMzUP-SAM-rel3-networks-cytoscape-ctrl.qmd](10-NAMzUP-SAM-rel3-networks-cytoscape-ctrl.qmd)
:::

## Subset networks by interesting GO-categories

### Prepare GO with descendants of categories of interest.


:::(.callout-caution}
This is important, otherwise only genes with direct annotation would be selected.
:::

```{r descendants}
# The category files from this chunk are universal
library(ontologyIndex)

# read obo file
go <- get_ontology("../../../input-data/go-uniprot/source-files/go.obo", extract_tags = "everything")

# get descendants of the Biological Process, Molecular Function and Cellular Component terms
dev_proc_descendants <- get_descendants(go, "GO:0032502") # developmental process
tran_fac_descendants <- get_descendants(go, "GO:0003700") # DNA-binding transcription factor activity
photosyn_descendants <- get_descendants(go, "GO:0015979") # photosynthesis
rhythm_proc_descendants <- get_descendants(go, "GO:0048511") # rhythmic process

# save to files, needed for graph drawing with go_plot.py
write(dev_proc_descendants, file = "../../../input-data/go-uniprot/dev-proc_descendants.txt")
write(tran_fac_descendants, file = "../../../input-data/go-uniprot/tran-fac_descendants.txt")
write(photosyn_descendants, file = "../../../input-data/go-uniprot/photosyn_descendants.txt")
write(rhythm_proc_descendants, file = "../../../input-data/go-uniprot/rhythm_proc_descendants.txt")
#write(cc_descendants, file = "GO_CC_descendants.txt")

# Check package Goplot
```

```{bash go check}
cd ../../../input-data/go-uniprot/

# DNA tran-fac activ
wr_hier.py GO:0003700 --dag source-files/go.obo --concise -o x
grep -Eo 'GO:[0-9]{7}' x | sort -u > x3
sort tran-fac_descendants.txt > x4

diff -s x3 x4
# Pliki x3 i x4 są identyczne

# developmental
wr_hier.py GO:0032502 --dag source-files/go.obo --concise -o x
grep -Eo 'GO:[0-9]{7}' x | sort -u > x3
sort dev-proc_descendants.txt > x4

diff -s x3 x4
# Pliki x3 i x4 są identyczne
```

### Extract from `NAM2GO_all_4making_pop`

```{bash descendants}
# The category files from this chunk are universal
# NAMs with selected GOs

cd ../../../input-data/go-uniprot/

# DNA transcription-factor activity
grep -Fwf tran-fac_descendants.txt NAM2GO_all_4making_pop | cut -f1 -d $'\t' | sort -u > genes-tran-reg

# Developmental process
grep -Fwf dev-proc_descendants.txt NAM2GO_all_4making_pop | cut -f1 -d $'\t' | sort -u > genes-developmental

# Photosynthesis
grep -Fwf photosyn_descendants.txt NAM2GO_all_4making_pop | cut -f1 -d $'\t' | sort -u > genes-photosyn

# Rhythmic process
grep -Fwf rhythm_proc_descendants.txt NAM2GO_all_4making_pop | cut -f1 -d $'\t' | sort -u > genes-rhythm-proc

# Header for networks for selected GOs
echo -e "uniprot1\tensembl1\tdesc1\tcluster1\tmember1\tuniprot2\tensembl2\tdesc2\tcluster2\tmember2\tneighborhood\tneighborhood_transferred\tfusion\tcooccurence\thomology\tcoexpression\tcoexpression_transferred\texperiments\texperiments_transferred\tdatabase\tdatabase_transferred\ttextmining\ttextmining_transferred\tcombined_score" > header-net-tab
```

```{bash tran-reg}

cd ../results/network/

for i in {1..10}; do
awk '
    NR==FNR {           # Czytamy pierwszy plik (genes-tran-reg)
        keep[$1] = 1;       # budujemy tablicę asocjacyjną – klucz = wartość
       next;             # przejdź do kolejnego rekordu (nie analizujemy dalej)
    }
    $2 in keep {         # Drugi plik (klastry); sprawdzamy kolumnę 2
        print               # wypisujemy cały wiersz, jeśli wartość jest w tablicy
    }
' ../../../../input-data/go-uniprot/genes-tran-reg cluster_${i}_400.txt > x
  cat header-net-tab x > cluster_${i}_400_tran-reg.txt;
done

## Remove files without data

files=($(ls cluster_*_400_tran-reg.txt))

for file in ${files[@]}; do
# Przykład: liczymy linie w pliku data.txt
linijek=$(wc -l < ${file})   # < – przekierowanie, dzięki czemu wc zwraca samą liczbę

if (( linijek < 2 )); then
rm ${file}
fi
done
```

```{bash dev-rel extract}
cd ../results/network/

for i in {1..10}; do
awk '
    NR==FNR {           # Czytamy pierwszy plik (genes-biol-reg)
        keep[$1] = 1;       # budujemy tablicę asocjacyjną – klucz = wartość
       next;             # przejdź do kolejnego rekordu (nie analizujemy dalej)
    }
    $2 in keep {         # Drugi plik (klastry); sprawdzamy kolumnę 2
        print               # wypisujemy cały wiersz, jeśli wartość jest w tablicy
    }
' ../../../../input-data/go-uniprot/genes-developmental cluster_${i}_400.txt > x
  cat header-net-tab x > cluster_${i}_400_dev.txt;
done

## Remove files without data

files=($(ls cluster_*_400_dev.txt))

for file in ${files[@]}; do
# Przykład: liczymy linie w pliku data.txt
linijek=$(wc -l < ${file})   # < – przekierowanie, dzięki czemu wc zwraca samą liczbę

if (( linijek < 2 )); then
rm ${file}
fi
done
```

```{bash rhythm extract}
cd ../results/network/

for i in {1..10}; do
awk '
    NR==FNR {           # Czytamy pierwszy plik (genes-biol-reg)
        keep[$1] = 1;       # budujemy tablicę asocjacyjną – klucz = wartość
       next;             # przejdź do kolejnego rekordu (nie analizujemy dalej)
    }
    $2 in keep {         # Drugi plik (klastry); sprawdzamy kolumnę 2
        print               # wypisujemy cały wiersz, jeśli wartość jest w tablicy
    }
' ../../../../input-data/go-uniprot/genes-rhythm-proc cluster_${i}_400.txt > x
  cat header-net-tab x > cluster_${i}_400_rhythm.txt;
done

## Remove files without data

files=($(ls cluster_*_400_rhythm.txt))

for file in ${files[@]}; do
# Przykład: liczymy linie w pliku data.txt
linijek=$(wc -l < ${file})   # < – przekierowanie, dzięki czemu wc zwraca samą liczbę

if (( linijek < 2 )); then
rm ${file}
fi
done
```

## Tu skończyłem - dalsze kategorie analogiznie niżej (DO POPRAWY)

```{bash biol-reg}
# select genes annotated with biological regulation term

cd ../results/network/

for i in {1..10}; do
awk '
    NR==FNR {           # Czytamy pierwszy plik (genes-biol-reg)
        keep[$1] = 1;       # budujemy tablicę asocjacyjną – klucz = wartość
       next;             # przejdź do kolejnego rekordu (nie analizujemy dalej)
    }
    $2 in keep {         # Drugi plik (klastry); sprawdzamy kolumnę 2
        print               # wypisujemy cały wiersz, jeśli wartość jest w tablicy
    }
' ../../../../input-data/go-uniprot/genes-biol-reg cluster_${i}_400.txt > x
  cat header-net-tab x > cluster_${i}_400_biol-reg.txt;
done

files=($(ls cluster_*_400_biol-reg.txt))

for file in ${files[@]}; do
# Przykład: liczymy linie w pliku data.txt
linijek=$(wc -l < ${file})   # < – przekierowanie, dzięki czemu wc zwraca samą liczbę

if (( linijek < 2 )); then
rm ${file}
fi
done
```

## References