---
title: "Heatmap visualization of WGCNA modules"
format: html
date: "2026-05-04"
bibliography: ../../../refs-stat-anal.bib
execute:
  fig-format: png
  fig.dpi: 300
---

## Rationale

This script is based on [instructions from DESeq2](https://bioconductor.org/packages//release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#heatmap-of-the-count-matrix).

The aim is to visualize genes from WGCNA modules.

Built with R version `r getRversion()`

## R Preliminaries

Environment cleaning and set-up

```{r setup, include = FALSE}
library(pheatmap)
library(DESeq2)
library(RColorBrewer)
library(paletteer)
library(limma)
```

## Load data

```{r load}
# counts after vsd
load('../rdata-saved/wgcna/vsd-mad.RDa')

# modules
load('../rdata-saved/wgcna/modules4go16.RDa')
```


## Sort

```{r sort}
coldata_srt = coldata.leaf[order(coldata.leaf$line, coldata.leaf$timepoint),]

vsd.counts_srt <- vsd.counts[, rownames(coldata_srt)]
```

```{r check}
# Check if files are in concert
all(rownames(coldata_srt) %in% colnames(vsd.counts_srt))
all(rownames(coldata_srt) == colnames(vsd.counts_srt))
```

## Split counts by module

```{r counts modules}
# lista tabelek z genami z modułów, posortowane malejąco
counts_genes <- lapply(names(genes_modules), function(name) {
  ids_present <- intersect(genes_modules[[name]], rownames(vsd.counts_srt))
  df <- as.data.frame(vsd.counts_srt[ids_present, , drop = FALSE])
  df$mean_expr <- rowMeans(df)
  df <- df[order(df$mean_expr, decreasing = TRUE), , drop = FALSE]
  df$mean_expr <- NULL
  df
})
names(counts_genes) <- names(genes_modules)

```


## Heatmaps

```{r heat}
png(filename = "grey60.png", width = 2000, height = 1200, res = 300)

# annotation
annotation_col <- data.frame(
  line = coldata_srt$line,
  time = coldata_srt$timepoint
)
mat <- counts_genes[['grey60']][1:50, ]
rownames(annotation_col) <- colnames(mat)


# Użycie palety kolorów

# colors
ann_colors <- list(
  line = setNames(brewer.pal(length(unique(coldata_srt$line)), "Set2"), unique(coldata_srt$line)),
  time = setNames(paletteer_d("grDevices::blues9", length(unique(coldata_srt$timepoint))), unique(coldata_srt$timepoint))
)

#anno = data.frame(group = coldata_srt$group)
pheatmap(counts_genes[['grey60']][1:50, ], cluster_cols = FALSE, cluster_rows = FALSE, show_rownames = FALSE, show_colnames = FALSE, annotation_col = annotation_col, annotation_colors = ann_colors, border_color = NA)
dev.off()
```

```{r func1}
heat_all <- function(mod_name) {
  module <- counts_genes[[mod_name]]
  filename <- paste0(mod_name, "_all.png")
  png(filename, width=2000, height=1200, res=300)
  
  # annotation
  annotation_col <- data.frame(
    line = coldata_srt$line,
    time = coldata_srt$timepoint
  )
  
  rownames(annotation_col) <- colnames(module)
  
  # Użycie palety kolorów
  
  # colors
  ann_colors <- list(
    line = setNames(brewer.pal(length(unique(coldata_srt$line)), "Set2"), unique(coldata_srt$line)),
    time = setNames(paletteer_d("grDevices::blues9", length(unique(coldata_srt$timepoint))), unique(coldata_srt$timepoint))
  )
  
  #anno = data.frame(group = coldata_srt$group)
  pheatmap(module[1:50, ], cluster_cols = FALSE, cluster_rows = FALSE, show_rownames = FALSE, show_colnames = FALSE, annotation_col = annotation_col, annotation_colors = ann_colors, border_color = NA)
  dev.off()
}

#lapply()
# heat_ave('white')

lapply(names(counts_genes), heat_all)
```

# Heatmaps for averaged data

```{r average}
# varMetadata - description of variables
metadata=data.frame(labelDescription=c("pooled sample", "letter", "tissue", "line", "hour of pooling", "sample number", "day", "replication", "condition", "id", "reads [million]", "million of reads binned", "chronological time point", "group"), row.names=colnames(coldata_srt))

# metadata
phenoData=new("AnnotatedDataFrame", data=as.data.frame(coldata_srt), varMetadata=metadata)
phenoData

## check
# sampleNames(phenoData)
# varMetadata(phenoData)
# head(pData(phenoData))

# make ExpressionSet
vsd.eset <- ExpressionSet(assayData=vsd.counts_srt, phenoData=phenoData)

# average reps
average = t(avereps(t(exprs(vsd.eset)), ID = vsd.eset$group)) # output is a matrix

# check
dim(average)
dim(vsd.eset)

# Make reduced coldata for plotting
coldata_ave <- coldata_srt[!duplicated(coldata_srt$group), ]
```

## Split averaged counts by module

```{r counts modules}
# lista tabelek z genami z modułów, posortowane malejąco
counts_genes_ave <- lapply(names(genes_modules), function(name) {
  ids_present <- intersect(genes_modules[[name]], rownames(average))
  df <- as.data.frame(average[ids_present, , drop = FALSE])
  df$mean_expr <- rowMeans(df)
  df <- df[order(df$mean_expr, decreasing = TRUE), , drop = FALSE]
  df$mean_expr <- NULL
  df
})
names(counts_genes_ave) <- names(genes_modules)

```

## Heatmaps for averaged data

```{r heat}
png(filename = "grey60ave.png", width = 2000, height = 1200, res = 300)

# annotation
annotation_col <- data.frame(
  line = coldata_ave$line,
  time = coldata_ave$timepoint
)

mat <- counts_genes_ave[['grey60']][1:50, ]
rownames(annotation_col) <- colnames(mat)

# Użycie palety kolorów

# colors
ann_colors <- list(
  line = setNames(brewer.pal(length(unique(coldata_ave$line)), "Set2"), unique(coldata_ave$line)),
  time = setNames(paletteer_d("grDevices::blues9", length(unique(coldata_ave$timepoint))), unique(coldata_ave$timepoint))
)

#anno = data.frame(group = coldata_ave$group)
pheatmap(counts_genes_ave[['grey60']][1:50, ], cluster_cols = FALSE, cluster_rows = FALSE, show_rownames = FALSE, show_colnames = FALSE, annotation_col = annotation_col, annotation_colors = ann_colors, border_color = NA)
dev.off()
```

## Function to draw heatmaps in one go

```{r func2}
heat_ave <- function(mod_name) {
  module <- counts_genes_ave[[mod_name]]
  filename <- paste0(mod_name, "_ave.png")
  png(filename, width=2000, height=1200, res=300)
  
  # annotation
  annotation_col <- data.frame(
    line = coldata_ave$line,
    time = coldata_ave$timepoint
  )

  rownames(annotation_col) <- colnames(module)
  
  # Użycie palety kolorów
  
  # colors
  ann_colors <- list(
    line = setNames(brewer.pal(length(unique(coldata_ave$line)), "Set2"), unique(coldata_ave$line)),
    time = setNames(paletteer_d("grDevices::blues9", length(unique(coldata_ave$timepoint))), unique(coldata_ave$timepoint))
  )
  
  #anno = data.frame(group = coldata_ave$group)
  pheatmap(module[1:50, ], cluster_cols = FALSE, cluster_rows = FALSE, show_rownames = FALSE, show_colnames = FALSE, annotation_col = annotation_col, annotation_colors = ann_colors, border_color = NA)
  dev.off()
}

#lapply()
# heat_ave('white')

lapply(names(counts_genes_ave), heat_ave)
```

## References
