---
title: "Visualization of WGCNA modules"
format: html
date: "2026-05-04"
bibliography: ../../../refs-stat-anal.bib
execute:
  fig-format: png
  fig.dpi: 300
---

https://peterlangfelder.com/2018/11/25/working-with-categorical-variables/



## Rationale

The aim is to visualize genes from WGCNA modules and compare modules.
Preliminary script - plots counts for provided module for all genes, not readable.
Better script is [02-wgcna-leaf-ctrl-line4-5genes_heat.qmd](02-wgcna-leaf-ctrl-line4-5genes_heat.qmd)

Built with R version `r getRversion()`

## R Preliminaries

Environment cleaning and set-up

```{r setup, include = FALSE}
library(pheatmap)
library(DESeq2)
```

## Load data

```{r load}
# wgcna results
load('cons16merged.RDa')

load('../rdata-saved/wgcna/modules4go16.RDa')
```

## Split counts by module

```{r counts modules}
# lista tabelek z genami z modułów, posortowane malejąco
counts_genes <- lapply(names(genes_modules), function(name) {
  ids_present <- intersect(genes_modules[[name]], rownames(vsd.counts))
  df <- as.data.frame(vsd.counts[ids_present, , drop = FALSE])
  df$mean_expr <- rowMeans(df)
  df <- df[order(df$mean_expr, decreasing = TRUE), , drop = FALSE]
  df$mean_expr <- NULL
  df
})
names(counts_genes) <- names(genes_modules)

```


## Heatmaps

```{r heat}
png(filename = "yellow.png", width = 2000, height = 1200, res = 300)
#anno = data.frame(group = coldata.leaf$group)
pheatmap(counts_genes[['yellow']], cluster_cols = FALSE, cluster_rows = FALSE, show_rownames = FALSE)#, annotation_col = anno)
dev.off()
```
## References
